---
title: "Wildfire Generated Ozone and Asthma Burden Assessment"
author: "Ryan_Gan"
date: "February 14, 2017"
output: html_document
---

## Purpose of Document

Contains code and results for the health impact assessment (aka burden assement) of child asthma emergency department (ED) visits that may be attributed to wildfire-generated ozone (WFO~3~).

The general HIA/BA formula is as follows:

$\Delta y = y_{o} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk$

We modified it to a daily estimate as our mean delta ozone for a state was on a given smoky day.
Daily HIA/BA formula:

$\Delta y = \frac{y_{o}}{365 days} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk$

We further modified it to account for the observed number of smoky days in Brey and Fisher for each state to get an overal period estimate.
Period HIA/BA formula:
$\Delta y = [\frac{y_{o}}{365 days} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk] \times n smoky days$

And finally, to calculate a proportion of events per 100,000 children at risk:
$\frac{\Delta y}{pop.atrisk} = (\frac{y_{o}}{365 days} \times[1-e^{(-\beta \times \Delta exposure)}]) \times n smoky days$

## Estimate Terms

*y~0~:* Baseline pediatric asthma ED visit rate. Came from CDC vital health statistics. Available for most states for multiple years. We pooled state estimates across years using meta-analysis to get pooled estimates and standard errors for each state.

*WFO~3~ and n Smoky Days:* Change in wildfire-generated ozone measured in 1 part per billion. As multiple sites used in Brey and Fischer were present in each state, we again used a meta-analysis to pool ozone estimates and produce a standard error. We averaged the smoky days within a site as Poisson distributions only need a mean (mean = variance). There were a couple (14 I think) monitoring sites that we were not able to assign a state, as the latitude and longitude were likely on coastal areas where the shapefile we used to attempt to find which state it was located in did not have enough resolution (e.g. I think there was a site on an island off the coast of Florida that wasn't assigned a state). I tried a couple things to avoid missingness, but in the interest of time, I just used the code as is.

*Pop at Risk:* This comes from Behavioral Risk Factor Surveillance System data from 2005 to 2014. Reported asthma prevalence (in counts) was available for most states, strata, and years in the study. We used meta-analysis to pool the estimate population at risk across years for each state and strata, as well as produce a standard error around this estimate.

*Beta:* This has been revised. Beta estimate and standard error now come from the Mar and Koenig 2009 paper. This estimate is child-specific (ages 0-18) of same day concentrations of ozone and ED visits. They found a risk ratio of 1.10 (95%CI: 1.01-1.19) for an ED visit for every 10 ppb increase in ozone, which were averaged across two study sites. We decided on this estimate for our HIA because it was specific to children, had a continous estimate of ozone, in the U.S., and on the same day (lag day 0). I calculated the beta estimate to be 0.0095 for a 1 ppb increase, with a standard error of 0.004. 


```{r call libraries, message = F, echo = F, warning=F}
library(tidyverse)
library(scales)
```



```{r read data, message = F, warning=F, echo = F}
daily_df <- read_csv("./data/mc_estimates/mc_daily.csv")
period_df <- read_csv("./data/mc_estimates/mc_period.csv")
prop_df <- read_csv("./data/mc_estimates/mc_prop_100k.csv")
state_ozone <- read_csv("./data/state_delta_o3.csv")
hia_input <- read_csv("./data/state_strata_hia_estimates.csv")
```


Estimates of ED visit burden in U.S. children with asthma that may be due to exposure to ozone associated with smoke from wildfires. 

## Maps of State Ozone Associated with Wildfire Smoke

### Daily \(\Delta \) Ozone Map 

Elevated ozone associated with smoke.

```{r ozone map, message = F, warning=F, echo = F, results='asis'}
# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(state_ozone, by = c("region" = "State")) %>% 
  filter(!is.na(delta_o3))

# summary(state_vals_map_df$delta_o3)

# ozone map
ozone_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = delta_o3), alpha = 0.8) +
  scale_fill_gradientn("EOAS", 
                       colours = c("#0a1d5c","#4B1248", "#F0C27B"),
                       values = rescale(c(-3, 0, 15)),
                       limits=c(-3, 15)) +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())


ozone_map

```

### Map of State Average Number of Smoke-Affected Days (2005-2014)

Smoke-affected days.

```{r smoky days, message = F, warning=F, echo = F, results='asis'}

# smky days map
smoke_days_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = n_smoky), alpha = 0.7) +
  scale_fill_gradientn("State Average Number of \nSmoke-Affected Days \n 2005-2014", 
                      colours = c('#cfffdb', '#018c65', '#0d1469'),
                      values = rescale(c(10, 30, 66)),
                      limits=c(10, 66)) +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())


smoke_days_map

```


## Asthma Burden: \(\Delta \) Emergency Department Visits 

Tables and graphs are for daily and period estimates for each strata of interest, by state.

## Continental U.S. Daily \(\Delta \) Emergency Department Visits

Estimates for the overall burden of asthma were the summed estimates calculated for each state. I've included the strata and state-specific estimate inputs and uncertainty (standard error) for the HIA formula at the very end of this document since it's large.



```{r daily US estimate, message = F, warning=F, echo = F, results='asis'}

# Daily estimates -----
# create a dataframe with all strata estimates and bind rows for 
# small multiples plot
daily_df2 <- daily_df %>% 
  filter(state == "state_sum") %>% 
  select(group, median, lower_bound, upper_bound)

# table ----
knitr::kable(daily_df2, caption = paste0("Table 1: Estimated average daily ED visit burden",
  " that may be due to wildfire-generated ozone"))

# preserve order of group
daily_df2$group <- reorder(daily_df2$group, daily_df2$median)

# plot ----
daily_us_plot <- ggplot(daily_df2, aes(x = group, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("Strata") +
  ylab(expression(paste(Delta, "y Emergency Department Visit"))) +
  coord_flip() +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

daily_us_plot

```


### State-Specific Daily \(\Delta \) Emergency Department Visits

Below are the state-specific figure by state. I didn't print the table because it would be large.

```{r daily table and line graph, message = F, warning=F, echo = F, results='asis'}

# Daily estimates -----
# create a dataframe with all strata estimates and bind rows for 
# small multiples plot
daily_df2 <- daily_df %>% 
  filter(state != "state_sum") %>% 
  mutate(state = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE))

# reorder state name to be descending
daily_df2$state <- factor(daily_df2$state, 
  levels = rev(unique(as.character(daily_df2$state))))
# preserve order of group
daily_df2$group <- factor(daily_df2$group, 
  levels = unique(as.character(daily_df2$group)))

# small multiples plot
daily_plot <- ggplot(daily_df2, aes(x = state, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("State") +
  ylab(expression(paste("Smoke-Affected Day ", Delta, "y Emergency Department Visit"))) +
  coord_flip() +
  facet_wrap(~group) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

daily_plot

```

### Daily Map

This is the small multiples map of the median daily \(\Delta \) ED visits estimated on an average smoky day over the study period of 2005-2014. White spaces incidate a state that was missing estimates for some term in the HIA formula. Therefore we were not able to estimate for those states. 

### Note:Alabama doesn't have estimates for Whites.

```{r daily map, message = F, warning=F, echo = F, results='asis'}
# I need to create a similar dataframe for small multiples as before, but I
# need to preserve the lowercase state names to merge
# small multiples plot
daily_df_map <- daily_df2 %>% 
  # I don't want group as there is a group in spatial map
  rename(strata = group) %>% 
    mutate(state = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\L\\2", 
                      state, perl=TRUE))

# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(daily_df_map, by = c("region" = "state")) %>% 
  mutate(missing = ifelse(is.na(median), 1, NA)) %>% 
  filter(!is.na(strata))

# preserve order of strata (groups)
state_vals_map_df$strata <- factor(state_vals_map_df$strata, 
  levels = unique(as.character(state_vals_map_df$strata)))

daily_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = median), alpha = 0.7) +
  scale_fill_gradient(expression("ED Visits"), low = '#2c7bb6', high = '#d7191c') + 
  xlab("Longitude") +
  ylab("Latitude") +
  facet_wrap(~strata) + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

daily_map

```

## Study Period (2005-2014) \(\Delta \) Emergency Department Visits

These estimates were the daily estimates multipled by the average number of smoky days observed in the specific state and then summed over the study period.

```{r period US estimate, message = F, warning=F, echo = F, results='asis'}

# Period estimates -----
# create a dataframe with all strata estimates and bind rows for 
# small multiples plot
period_df2 <- period_df %>% 
  filter(state == "state_sum") %>% 
  select(group, median,lower_bound, upper_bound)

# table ----
knitr::kable(period_df2, caption = paste0("Table 2: Estimated period ED visit burden",
  " that may be due to wildfire-generated ozone"))

# preserve order of group
period_df2$group <- reorder(period_df2$group, period_df2$median)

# plot ----
period_us_plot <- ggplot(period_df2, aes(x = group, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("Strata") +
  ylab(expression(paste(Delta, " Emergency Department Visit"))) +
  coord_flip() +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

period_us_plot

```

### State-Specific Period \(\Delta \) Emergency Department Visits

These estimates were the daily estimates multipled by the average number of smoky days observed in the specific state over the study period.

```{r period line graph, message = F, warning=F, echo = F, results='asis'}
# Period estimates ----
# small multiples plot
period_df2 <- period_df %>% 
  filter(state != "state_sum") %>% 
   # capitalize first letter of each state
  mutate(state = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE))

# reorder state name to be descending
period_df2$state <- factor(period_df2$state, 
  levels = rev(unique(as.character(period_df2$state))))
# preserve order of group
period_df2$group <- factor(period_df2$group, 
  levels = unique(as.character(period_df2$group)))

# plot
period_plot <- ggplot(period_df2, aes(x = state, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("State") +
  ylab(expression(paste(Delta, " Emergency Department Visit"))) +
  coord_flip() +
  facet_wrap(~group) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

period_plot

```

### Period Map

This is the small multiples map of the median overall \(\Delta \) ED visits estimated on an average smoky day over the study period of 2005-2014. White spaces incidate a state that was missing estimates for some term in the HIA formula. Therefore were were not able to estimate for those states. 


```{r period map, message = F, warning=F, echo = F, results='asis'}
# I need to create a similar dataframe for small multiples as before, but I
# need to preserve the lowercase state names to merge
# small multiples plot
period_df_map <- period_df2 %>% 
  rename(strata = group) %>% 
    mutate(state = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\L\\2", 
                      state, perl=TRUE))

# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(period_df_map, by = c("region" = "state")) %>% 
  mutate(missing = ifelse(is.na(median), 1, NA)) %>% 
  filter(!is.na(strata))

# preserve order of strata (groups)
state_vals_map_df$strata <- factor(state_vals_map_df$strata, 
  levels = unique(as.character(state_vals_map_df$strata)))

period_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = median), alpha = 0.7) +
  scale_fill_gradient(expression("ED Visits"), low = '#2c7bb6', high = '#d7191c') + 
  xlab("Longitude") +
  ylab("Latitude") +
  facet_wrap(~strata) + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

period_map

```

## \(\Delta \) Emergency Department Visits over Population at Risk Maps

I did some simulations in a markdown file to test out this concept and I believe it can be thought of in a similar way to incidence proportion. It's actually closest in estimation to the attributable risk differece between exposed and unexposed. The HIA/BA formula can be modified by simply removing the term for population at risk. The calculated proportion is very small, so I think we'll multiply it by 100,000 children (like Jake originally did). I don't know the best way to phrase it, but I would probably call this something like estimated number of emergency department visits per 100,000 children with asthma that may attributed to WFO~3~. I did not estimate proportion every day, but only over the entire period (wouldn't be too hard to add daily though). 

### Estimated ED visit per 100k children 

Table and map.

### Table

There are some negative proportions, which shouldn't happen. But I think it's a result of not including population at risk in the HIA. However, to compare across groups/states accounting for population size, I think it works.  

```{r table prop, message = F, warning=F, echo = F, results='asis'}

prop_df2 <- prop_df %>% 
  filter(state == "state_sum") %>% 
  select(group, median, lower_bound, upper_bound, median_100k, lower_bound_100k,
         upper_bound_100k)

# table ----
knitr::kable(prop_df2, caption = paste0("Table 3: Estimated ED visit per 100k children",
  " that may be due to wildfire-generated ozone"))
```

### Proportion Map

Unlike the table, I set all values < 0 to 0 to avoid confusion.

```{r proportion map, message = F, warning=F, echo = F, results='asis'}
# I need to create a similar dataframe for small multiples as before, but I
# need to preserve the lowercase state names to merge
# small multiples plot
prop_df_map <- prop_df %>% 
  filter(state != "state_sum") %>% 
  # I don't want group as there is a group in spatial map
  rename(strata = group) %>% 
  mutate(median_100k = ifelse(median_100k < 0, 0, median_100k))


# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(prop_df_map, by = c("region" = "state")) %>% 
  mutate(missing = ifelse(is.na(median), 1, NA)) %>% 
  filter(!is.na(strata))

# preserve order of strata (groups)
state_vals_map_df$strata <- factor(state_vals_map_df$strata, 
  levels = unique(as.character(state_vals_map_df$strata)))

prop_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = median_100k), alpha = 0.7) +
  scale_fill_gradient(expression("ED Visits per \n100k Children \nwith Asthma"), 
                      low = '#2c7bb6', high = '#d7191c') + 
  xlab("Longitude") +
  ylab("Latitude") +
  facet_wrap(~strata) + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

prop_map


```

Output data as it was used in the document so MC results are the same.

### HIA Input Terms

```{r hia terms,  message = F, warning=F, echo = F, results='asis'}
knitr::kable(hia_input, caption = paste0("Table 4: HIA term input and standard error",
             " for each state and strata"))
```