---
title: "Wildfire Generated Ozone and Asthma Burden Assessment"
author: "Ryan_Gan"
date: "February 14, 2017"
output: html_document
---

## Purpose of Document

Contains code and results for the health impact assessment (aka burden assement) of child asthma emergency department (ED) visits that may be attributed to wildfire-generated ozone (WFO~3~).

The general HIA/BA formula is as follows:

$\Delta y = y_{o} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk$

We modified it to a daily estimate as our mean delta ozone for a state was on a given smoky day.
Daily HIA/BA formula:

$\Delta y = \frac{y_{o}}{365 days} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk$

We further modified it to account for the observed number of smoky days in Brey and Fisher for each state to get an overal period estimate.
Period HIA/BA formula:
$\Delta y = [\frac{y_{o}}{365 days} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk] \times n smoky days$

And finally, to calculate a proportion of events per 100,000 children at risk:
$\frac{\Delta y}{pop.atrisk} = (\frac{y_{o}}{365 days} \times[1-e^{(-\beta \times \Delta exposure)}]) \times n smoky days$


```{r call libraries, message = F, echo = F, warning=F}
library(tidyverse)
#library(ggmap)
```

```{r data read, message = F, echo = F,  warning = F}

# read smoky days df
sd_path <- paste0('./data/smoky_days.csv')
smoky_days <- read_csv(sd_path) 

# start with one loop first of the total state-specific estimates
total_path <- paste0('./data/Total.csv')
# import csv and make sure variables as numeric like they should be
total <- read_csv(total_path, col_types = "cdddddddd", na = "NA") 

# read in african american df
black_path <- paste0('./data/Black.csv')
black <- read_csv(black_path, col_types = "cdddddddd", na = "NA")


# hispanic
hisp_path <- paste0('./data/Hispanic.csv')
hisp <- read_csv(hisp_path, col_types = "cdddddddd", na = "NA") 


# white
white_path <- paste0('./data/White.csv')
white <- read_csv(white_path, col_types = "cdddddddd", na = "NA") 


# male
m_path <- paste0('./data/Male.csv')
male <- read_csv(m_path, col_types = "cdddddddd", na = "NA") 

# female
f_path <- paste0('./data/Female.csv')
female <- read_csv(f_path, col_types = "cdddddddd", na = "NA")


# create a list of dataframes to cycle through
df_list <- list(total=total, male=male, female=female, black=black, 
                hisp=hisp, white=white)

df_name <- c('total', 'male', 'female', 'black', 'hisp', 'white')

# join smokey days df to each dataframe
for(q in 1:length(df_list)){

  # loop through df list and join smokey days variable to strata dfs
  data_join_df <- df_list[[q]] %>% 
    full_join(smoky_days, by = "State") %>% filter(complete.cases(.))
  # rename original df lists to contain smoky days val
  assign(names(df_list[q]), data_join_df)
}

# re-create list of dataframes to cycle through now that it has smoke days
df_list <- list(total=total, male=male, female=female, black=black, 
                hisp=hisp, white=white)
# check out first element of list to make sure smoky days variable is there
# glimpse(df_list[2]) # it is

```

Estimates of ED visit burden in U.S. children with asthma that may be due to exposure to ozone generated from wildfires. Beta estimate and standard error now come from the Mar and Koenig 2009 paper. This estimate is child-specific (ages 0-18) of same day concentrations of ozone and ED visits. They found a risk ratio of 1.10 (95%CI: 1.01-1.19) for an ED visit for every 10 ppb increase in ozone, which were averaged across two study sites. We decided on this estimate for our HIA because it was specific to children, had a continous estimate of ozone, in the U.S., and on the same day (lag day 0). I calculated the beta estimate to be 0.0095 for a 1 ppb increase, with a standard error of 0.004. 

Note: I will need to remove beta and SE from datasets. Yo and beta are now simulated outside the MC so that we draw from the same distribution each time.

```{r monte carlo estimate, message = F, echo = F, warning = F}

# Loop 
# Set seed  to use so sample distributions and resampling is somewhat similar
sim_seed <- 1234
# number of interations for HIA MC and simulated distribution sample size
n <- 10000

# simulate beta distribution as this estimate applies across all strata
# using the same beta estimate for all simulations
beta <- 0.0095
beta_se <- 0.0042
  
# distribution of beta
set.seed(sim_seed)
beta_distribution <- rnorm(n, mean = beta, sd = beta_se)

# check on beta distribution
# ggplot(data = as.data.frame(beta_distribution), aes(x=beta_distribution))+ 
#   geom_density()
# 
# # check against original estimates
# exp(as.numeric(quantile(beta_distribution, 0.50, na.rm = T))*10)
# # 2.5% lower bound
# exp(as.numeric(quantile(beta_distribution, 0.025, na.rm = T))*10)
# # 97.5 % upper bound
# exp(as.numeric(quantile(beta_distribution, 0.975, na.rm = T))*10)


# strata loop ---
for(k in 1:length(df_list)){

  # set dataframe to loop through  
  # Note: changing name 'data_frame' to 'df_to_loop' 
  # since it's also a function in dplyr
  df_to_loop <- data.frame(df_list[k]) 
  df_name <- names(df_list[k])

  # yo and beta as well as their std errors do not change
  # simulate the distributions outside the loop
  # output the value for alabama since they are the same for each state
  # these estimate comes from a meta analysis of the estimate and prev.
  yo_total <- df_to_loop[1,2] 
  yo_se_total <- df_to_loop[1,3] 

  yo_distribution <- rnorm(n, mean = yo_total, sd = yo_se_total)

  # yo check  
  # ggplot(data = as.data.frame(yo_distribution), aes(x=yo_distribution))+ 
  #   geom_density()
  
  # create an empty matrix to populate with state specific median and bounds
  hia_matrix <- matrix(data=NA , nrow(df_to_loop), 5) 
  colnames(hia_matrix) <- c("state", "median", "lower_bound", "upper_bound",
                            "group")
  
  # convert hia_matrix to dataframe
  hia_prop_df <- hia_df <- hia_daily_df <- data.frame(hia_matrix)
  # duplicate empty df to fill with daily estimates
  
  # assign a group name
  hia_prop_df[,5] <- hia_df[,5] <- hia_daily_df[,5] <- df_name
  
  # start 2nd loop to run MC for each state ----
  for(i in 1:nrow(df_to_loop)){

    # feed state name in to hia dataframe
    hia_df[[i, 1]] <- as.character(df_to_loop[i, 1])
    hia_daily_df[[i, 1]] <- as.character(df_to_loop[i, 1])
    hia_prop_df[[i, 1]] <- as.character(df_to_loop[i, 1])
    
    # Now I need to simulate the distributions for population at risk and 
    # output the state-specific pop_at_risk values
    state_par_val <- df_to_loop[i, 4] 
    state_par_se <- df_to_loop[i, 5] 
    # create state-specific pop at risk distribution
    state_par_distribution <- rnorm(n, mean = state_par_val, sd = state_par_se)
    
    # output the state-specific delta ozone values
    state_delta_o3 <- df_to_loop[i, 6]
    state_do3_se <- df_to_loop[i, 7]
    # create state-specific o3 distribution
    state_o3_distribution <- rnorm(n, mean = state_delta_o3, sd = state_do3_se)
    
    # output the state-specific n smoke days value
    state_smk_d_n <- df_to_loop[i, 10]
    # create state_specific smoke days Poisson dist
    state_smoky_days_distribution <- rpois(n, state_smk_d_n)
    
    # make empty vectors to fill with change in y
    # daily 
    delta_y_daily <- vector("double", length = n)
    # period
    delta_y <- vector("double", length = n)
    # period proportion
    prop_y <- vector("double", length = n)
    
    # beginning of inner loop to estimate n interations of HIA formula ----
    for(j in 1:n) {
      # take one random value with replacement from the ED_rate distribution
      # divide by 365 to estimate daily rate
      est_yo <- (sample(yo_distribution, 1, replace = T))/365
      # take one random value from the beta distribution with replacement
      est_beta <- sample(beta_distribution, 1, replace = T)
      # take one random value from the ozone distribution with replacement
      est_delta_o3 <- sample(state_o3_distribution, 1, replace = T)
      # take one random value from the population at risk distribution with replacement
      pop_at_risk <- sample(state_par_distribution, 1, replace = T)
      #take a random value from the poisson distribution of smoky days
      est_smoky_days <- sample(state_smoky_days_distribution, 1, replace = T)
      
      
      # apply the HIA formula with the randomly selected values above
      # and save the estimate delta y in the vector, n times to
      # fill in the state-specific median, 2.5% lower bound, and 97.5% upper bound
      # from the monte carlo vector
      
      # average daily estimate
      delta_y_daily[[j]] <- (est_yo*(1-exp((-est_beta)*(est_delta_o3)))*pop_at_risk)
      # average period estimate
      delta_y[[j]] <- (est_yo*(1-exp((-est_beta)*(est_delta_o3)))*pop_at_risk)*est_smoky_days
      
      # adding a new formula to calculate the delta y proportion 
      # that may be due to wfo3 
      prop_y[[j]] <- (est_yo*(1-exp((-est_beta)*(est_delta_o3))))*est_smoky_days

    } # end of inner loop of HIA estimate
    
    # now calculate the median value of delta y and fill in dataframes
    # daily estimates ----
    hia_daily_df[[i, 2]] <- as.numeric(quantile(delta_y_daily, 0.50, na.rm = T))
    # 2.5% lower bound
    hia_daily_df[[i, 3]]<- as.numeric(quantile(delta_y_daily, 0.025, na.rm = T))
    # 97.5 % upper bound
    hia_daily_df[[i, 4]] <- as.numeric(quantile(delta_y_daily, 0.975, na.rm = T))
    
    # smoke period estimates ----
    hia_df[[i, 2]] <- as.numeric(quantile(delta_y, 0.50, na.rm = T))
    # 2.5% lower bound
    hia_df[[i, 3]]<- as.numeric(quantile(delta_y, 0.025, na.rm = T))
    # 97.5 % upper bound
    hia_df[[i, 4]] <- as.numeric(quantile(delta_y, 0.975, na.rm = T))
    
    # smoke period for proportion estimates ----
    hia_prop_df[[i, 2]] <- as.numeric(quantile(prop_y, 0.50, na.rm = T))
    # 2.5% lower bound
    hia_prop_df[[i, 3]]<- as.numeric(quantile(prop_y, 0.025, na.rm = T))
    # 97.5 % upper bound
    hia_prop_df[[i, 4]] <- as.numeric(quantile(prop_y, 0.975, na.rm = T))
    
  } # end state specific
  
# U.S. estimates for daily and period
# daily ----
# add a final sum of each column to get a U.S. estimate
us_estimate <- cbind("state_sum", sum(hia_daily_df[,2]), sum(hia_daily_df[,3]),
  sum(hia_daily_df[,4]), df_name) 
# assign same column name
colnames(us_estimate) <- colnames(hia_daily_df)
# bind US estimate to full dataframe
hia_daily_df <- rbind(hia_daily_df, us_estimate)
  
# period ----
# add a final sum of each column to get a U.S. estimate
us_estimate <- cbind("state_sum", sum(hia_df[,2]), sum(hia_df[,3]),
  sum(hia_df[,4]), df_name) 
# assign same column name
colnames(us_estimate) <- colnames(hia_df)
# bind US estimate to full dataframe
hia_df <- rbind(hia_df, us_estimate)
  
# proportion ----
# add a final sum of each column to get a U.S. estimate
us_estimate <- cbind("state_sum", sum(hia_prop_df[,2]), sum(hia_prop_df[,3]),
  sum(hia_prop_df[,4]), df_name) 
# assign same column name
colnames(us_estimate) <- colnames(hia_prop_df)
# bind US estimate to full dataframe
hia_prop_df <- rbind(hia_prop_df, us_estimate)

# create environment dataframe with name of original dataset hia is based off
# period estimates
hia_p_name <- paste(df_name, 'hia_period', sep = '_')
assign(hia_p_name, hia_df)
# write permanent dataset
#write_path <- paste('./data/mc_estimates/', hia_p_name, '.csv', sep = '')
#write_csv(hia_df, write_path)
# daily estimates
hia_d_name <- paste(df_name, 'hia_daily', sep = '_')
assign(hia_d_name, hia_daily_df)
# write permanent dataset
#write_path <- paste('./data/mc_estimates/', hia_d_name, '.csv', sep = '')
#write_csv(hia_daily_df, write_path)

# period proportion estimates
hia_pr_name <- paste(df_name, 'hia_prop', sep = '_')
assign(hia_pr_name, hia_prop_df)
# write permanent dataset
#write_path <- paste('./data/mc_estimates/', hia_pr_name, '.csv', sep = '')
#write_csv(hia_prop_df, write_path)


}

# end dataframe specific loop
# check to make sure delta ozone and beta are the same units 

# Important Note ----
# rerunning this code will in the future will not produce the exact same
# median and bound values because of the simulated distributions and random replacement
# will not be the same each time, but the values should be very close.


```


## Asthma Burden: \(\Delta \) Emergency Department Visits 

Tables and graphs are for daily and period estimates for each strata of interest, by state.

## Continental U.S. Daily \(\Delta \) Emergency Department Visits

Estimates for the overall burden of asthma were the summed estimates calculated for each state.

```{r daily US estimate, message = F, warning=F, echo = F, results='asis'}

# Daily estimates -----
# create a dataframe with all strata estimates and bind rows for 
# small multiples plot
daily_df <- rbind(total_hia_daily, female_hia_daily, male_hia_daily,
                  white_hia_daily, black_hia_daily, hisp_hia_daily) %>% 
  filter(state == "state_sum") %>% 
  # capitalize first letter of each state
  mutate(median = as.numeric(median),
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group) %>% 
  rename(group = group2) %>% 
  select(group, median, lower_bound, upper_bound)

# table ----
knitr::kable(daily_df, caption = paste0("Table 1: Estimated average daily ED visit burden",
  " that may be due to wildfire-generated ozone"))

# preserve order of group
daily_df$group <- reorder(daily_df$group, daily_df$median)

# plot ----
daily_us_plot <- ggplot(daily_df, aes(x = group, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("Strata") +
  ylab(expression(paste(Delta, "y Emergency Department Visit"))) +
  coord_flip() +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

daily_us_plot

```


### State-Specific Daily \(\Delta \) Emergency Department Visits

Below are the state-specific figure by state. I didn't print the table because it would be large.

```{r daily table and line graph, message = F, warning=F, echo = F, results='asis'}

# Daily estimates -----
# create a dataframe with all strata estimates and bind rows for 
# small multiples plot
daily_df <- rbind(total_hia_daily, female_hia_daily, male_hia_daily,
                  white_hia_daily, black_hia_daily, hisp_hia_daily) %>% 
  filter(state != "state_sum") %>% 
  # capitalize first letter of each state
  mutate(state = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE),
         median = as.numeric(median),
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group) %>% 
  rename(group = group2)

# reorder state name to be descending
daily_df$state <- factor(daily_df$state, 
  levels = rev(unique(as.character(daily_df$state))))
# preserve order of group
daily_df$group <- factor(daily_df$group, 
  levels = unique(as.character(daily_df$group)))

# small multiples plot
daily_plot <- ggplot(daily_df, aes(x = state, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("State") +
  ylab(expression(paste(Delta, "y Emergency Department Visit"))) +
  coord_flip() +
  facet_wrap(~group) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

daily_plot

```

### Daily Map

This is the small multiples map of the median daily \(\Delta \) ED visits estimated on an average smoky day over the study period of 2005-2014. White spaces incidate a state that was missing estimates for some term in the HIA formula. Therefore we were not able to estimate for those states. 

### Note:Alabama doesn't have estimates for Whites.

```{r daily map, message = F, warning=F, echo = F, results='asis'}
# I need to create a similar dataframe for small multiples as before, but I
# need to preserve the lowercase state names to merge
# small multiples plot
daily_df <- rbind(total_hia_daily, female_hia_daily, male_hia_daily,
                  white_hia_daily, black_hia_daily, hisp_hia_daily) %>% 
  filter(state != "state_sum") %>% 
  # capitalize first letter of each state
  mutate(state2 = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE),
         median = as.numeric(median),
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group, -lower_bound, -upper_bound) %>% 
  # I don't want group as there is a group in spatial map
  rename(strata = group2) 

# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(daily_df, by = c("region" = "state")) %>% 
  mutate(missing = ifelse(is.na(median), 1, NA)) %>% 
  filter(!is.na(strata))

# preserve order of strata (groups)
state_vals_map_df$strata <- factor(state_vals_map_df$strata, 
  levels = unique(as.character(state_vals_map_df$strata)))

daily_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = median), alpha = 0.7) +
  scale_fill_gradient(expression("ED Visits"), low = '#2c7bb6', high = '#d7191c', 
                      guide = guide_legend(title.position = "top", order = 1)) +
  xlab("Longitude") +
  ylab("Latitude") +
  facet_wrap(~strata) + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

daily_map

```

## Study Period (2005-2014) \(\Delta \) Emergency Department Visits

These estimates were the daily estimates multipled by the average number of smoky days observed in the specific state and then summed over the study period.

```{r period US estimate, message = F, warning=F, echo = F, results='asis'}

# Period estimates -----
# create a dataframe with all strata estimates and bind rows for 
# small multiples plot
period_df <- rbind(total_hia_period, female_hia_period, male_hia_period,
                  white_hia_period, black_hia_period, hisp_hia_period) %>% 
  filter(state == "state_sum") %>% 
  # capitalize first letter of each state
  mutate(median = as.numeric(median),
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group) %>% 
  rename(group = group2) %>% 
  select(group, median, lower_bound, upper_bound)

# table ----
knitr::kable(period_df, caption = paste0("Table 2: Estimated period ED visit burden",
  " that may be due to wildfire-generated ozone"))

# preserve order of group
period_df$group <- reorder(period_df$group, period_df$median)

# plot ----
period_us_plot <- ggplot(period_df, aes(x = group, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("Strata") +
  ylab(expression(paste(Delta, " Emergency Department Visit"))) +
  coord_flip() +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

period_us_plot

```

### State-Specific Period \(\Delta \) Emergency Department Visits

These estimates were the daily estimates multipled by the average number of smoky days observed in the specific state over the study period.

```{r period line graph, message = F, warning=F, echo = F, results='asis'}
# Period estimates ----
# small multiples plot
period_df <- rbind(total_hia_period, female_hia_period, male_hia_period,
                  white_hia_period, black_hia_period, hisp_hia_period) %>% 
  filter(state != "state_sum") %>% 
  # capitalize first letter of each state
  mutate(state = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE),
         median = as.numeric(median),
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group) %>% 
  rename(group = group2)

# reorder state name to be descending
period_df$state <- factor(period_df$state, 
  levels = rev(unique(as.character(period_df$state))))
# preserve order of group
period_df$group <- factor(period_df$group, 
  levels = unique(as.character(period_df$group)))

# plot
period_plot <- ggplot(period_df, aes(x = state, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound)) +
  geom_hline(yintercept = 0, linetype = 'longdash', color = 'red') +
  xlab("State") +
  ylab(expression(paste(Delta, " Emergency Department Visit"))) +
  coord_flip() +
  facet_wrap(~group) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

period_plot

```

### Period Map

This is the small multiples map of the median overall \(\Delta \) ED visits estimated on an average smoky day over the study period of 2005-2014. White spaces incidate a state that was missing estimates for some term in the HIA formula. Therefore were were not able to estimate for those states. 


```{r period map, message = F, warning=F, echo = F, results='asis'}
# I need to create a similar dataframe for small multiples as before, but I
# need to preserve the lowercase state names to merge
# small multiples plot
period_df <- rbind(total_hia_period, female_hia_period, male_hia_period,
                  white_hia_period, black_hia_period, hisp_hia_period) %>% 
  filter(state != "state_sum") %>% 
  # capitalize first letter of each state
  mutate(state2 = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE),
         median = as.numeric(median),
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group, -lower_bound, -upper_bound) %>% 
  # I don't want group as there is a group in spatial map
  rename(strata = group2) 

# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(period_df, by = c("region" = "state")) %>% 
  mutate(missing = ifelse(is.na(median), 1, NA)) %>% 
  filter(!is.na(strata))

# preserve order of strata (groups)
state_vals_map_df$strata <- factor(state_vals_map_df$strata, 
  levels = unique(as.character(state_vals_map_df$strata)))

period_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = median), alpha = 0.7) +
  scale_fill_gradient(expression("ED Visits"), low = '#2c7bb6', high = '#d7191c', 
                      guide = guide_legend(title.position = "top", order = 1)) +
  xlab("Longitude") +
  ylab("Latitude") +
  facet_wrap(~strata) + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

period_map

```

## \(\Delta \) Emergency Department Visits over Population at Risk Maps

I did some simulations in a markdown file to test out this concept and I believe it can be thought of in a similar way to incidence proportion. It's actually closest in estimation to the attributable risk differece between exposed and unexposed. The HIA/BA formula can be modified by simply removing the term for population at risk. The calculated proportion is very small, so I think we'll multiply it by 100,000 children (like Jake originally did). I don't know the best way to phrase it, but I would probably call this something like estimated number of emergency department visits per 100,000 children with asthma that may attributed to WFO~3~. I did not estimate proportion every day, but only over the entire period (wouldn't be too hard to add daily though). 

### Small Multiples Per 100,000 Map 

```{r per 100k map, message = F, warning=F, echo = F, results='asis'}
# I need to create a similar dataframe for small multiples as before, but I
# need to preserve the lowercase state names to merge
# small multiples plot
prop_df <- rbind(total_hia_prop, female_hia_prop, male_hia_prop,
                  white_hia_prop, black_hia_prop, hisp_hia_prop) %>% 
  filter(state != "state_sum") %>% 
  # capitalize first letter of each state
  mutate(state2 = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2", 
                      state, perl=TRUE),
         median = as.numeric(median)*100000,
         lower_bound = as.numeric(lower_bound),
         upper_bound = as.numeric(upper_bound),
         group2 = ifelse(group == "total", "Total",
                  ifelse(group == "female", "Female",
                  ifelse(group == "male", "Male",
                  ifelse(group == "white", "White",
                  ifelse(group == "black", "Black",
                  ifelse(group == "hisp", "Hispanic", NA))))))) %>% 
  select(-group, -lower_bound, -upper_bound) %>% 
  # I don't want group as there is a group in spatial map
  rename(strata = group2) 

# create maps using ggplot2 ----
state_map_df <- map_data("state")

# merge median values to statemap
state_vals_map_df <- state_map_df %>% 
  full_join(prop_df, by = c("region" = "state")) %>% 
  mutate(missing = ifelse(is.na(median), 1, NA)) %>% 
  filter(!is.na(strata))

# preserve order of strata (groups)
state_vals_map_df$strata <- factor(state_vals_map_df$strata, 
  levels = unique(as.character(state_vals_map_df$strata)))

prop_map <- ggplot(state_map_df, aes(x = long, y = lat, group=group)) +
  # outline of states
  geom_path() +
  # fill with median value of delta y
  geom_polygon(data = state_vals_map_df, aes(fill = median), alpha = 0.7) +
  scale_fill_gradient(expression("ED Visits per \n100k Children \nwith Asthma"), 
                      low = '#2c7bb6', high = '#d7191c', 
                      guide = guide_legend(title.position = "top", order = 1)) +
  xlab("Longitude") +
  ylab("Latitude") +
  facet_wrap(~strata) + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"))

prop_map

```




### Notes: 
Colors on maps could be changed as well as the text size on the line plots (State names are hard to read). Also, we should do a final check of all the inputs going in to the Monte Carlo. Otherwise, things look in order.