---
title: "What's the point of an HIA?"
author: "Ryan_Gan"
date: "February 16, 2017"
output: html_document
---

```{r library, message=F}
library(tidyverse)
library(broom)
```

## Document purpose

To explore how the Health Impact Assessment/Burden Assessment formula is supposed to work. In theory, it can be used to estimate the impact of a pollutant/policy on population health in the form of an estimated count of outcomes that may be due to that pollutant/policy (assuming a causal relationship) using information from existing sources and without having to undertake a study to specifically answer the question.

Going to start with a simulated 'true' association for a binary outcome.

```{r population simulation, message=F}
# simulation population data ----
set.seed(987)
pop_data <- data_frame(x = rep(seq(from = 1, to = 100, by =1), 100)) %>% 
  arrange(x) %>%  # sort x by ascending values
  cbind(y = rep(seq(from = 1, to = 100, by =1), 100)) %>% 
  # create randomly assigned proportion of disease
  mutate(exposure = rbinom(10000, size = 1, prob = 0.5),
         exp_yn = ifelse(exposure == 1, "Yes", "No"))


# finding the formula of the baseline disease probability I want
# 1/(1+exp(-(-3.6 + 0.69))) 
# logit relationship between dis and exp
logit_form = -3.33 + 0.69*pop_data$exposure # linear combination with a bias
# define probability of disease given formula
pr = 1/(1+exp(-logit_form))   

# simulate disease
pop_data <- pop_data %>% 
  cbind(disease = rbinom(10000, size = 1, prob = pr)) %>% 
  mutate(dis_yn = ifelse(disease == 1, "Yes", "No"),
         # make 4 category exposure/disease variable
         exp_dis = as.factor(
                   ifelse(exposure == 0 & disease == 0, "Exp = N, Dis = N",
                   ifelse(exposure == 0 & disease == 1, "Exp = N, Dis = Y",
                   ifelse(exposure == 1 & disease == 0, "Exp = Y, Dis = N",
                   ifelse(exposure == 1 & disease == 1, "Exp = Y, Dis = Y", 
                          NA))))))

```

In the simulation above, I've created a binary exposure with a distribution in the population of roughly 50%. I then defined the exposure/response relationship to be an odds ratio of ~2.00 (or a beta of 0.69). Also, the baseline proportion of disease in this population is ~0.05 or 5%. Below are our estimates of 'truth'.

Contigency table of simulated data by exposure/disease status.

```{r 2x2 table, message=F}
xtabs(~exposure + disease, pop_data)
```

Proportion of disease in population (analagous to y~o~ in formula).

```{r incidence proportion of disease, message = F}
mean(pop_data$disease)
```

Odds ratio and \beta for the relationshi (same \beta used in formula). Estimates just as we defined.

```{r beta calcs, message = F}
# hand calc of odds ratio 
(340/4629)/(161/4870)
# and beta
log((340/4629)/(161/4870))

# logistic regression beta
logit_mod <- tidy(glm(disease~exposure, data = pop_data, family = "binomial"(link="logit")))
logit_mod

# odds ratio
exp(logit_mod[2,2])

# relative risk ratio
# logistic regression beta
riskratio_mod <- tidy(glm(disease~exposure, data = pop_data, family = "poisson"(link="log")))
riskratio_mod

# risk ratio
exp(riskratio_mod[2,2])
```

So now we know the 'true' odds/risk ratio and \beta. We can also calculate the 'true' attributable risk difference and even a count of the outcome that is due to exposure.

```{r attrib risk and count, message=F}
# hand calc risk difference
(340/(340+4629))-(161/(161+4870))

# risk diff
# logistic regression beta
risk_diff_mod <- tidy(glm(disease~exposure, data = pop_data, 
                          family = "binomial"(link="identity")))
risk_diff_mod

# count difference (attributed to exposure)
340-161

```

“The general health impact/burden assesment formula is as follows:

$\Delta y = y_{o} \times[1-e^{(-\beta \times \Delta exposure)}] \times pop. at risk$

Let's see if we get the same answers as the 'truth' using the formula. 
Let $y_{o} = 0.0501$; $-\beta = -0.7983$; $\Delta exposure = 1$; $ pop. at risk = 10,000$

```{r hia estimate using vals, message=F}
# hia estimate
yo <- 0.0501*(1-exp(-0.7983*1))*10000 
yo
# as a reminder, here is the actual count  
340-161
```

Let's see if a Monte-Carlo estimate can capture the 'true' estimate in it's 95% CI bounds. For simplicity, I"m just going to calculate a distribution around baseline population and beta.

```{r mc estimate, message = F}
# intercept only mod to get SE around incident proportion
yo_mod <- tidy(glm(disease~ 1, data = pop_data, 
               family = "binomial"(link="identity")))
yo_mod
yo_est <- yo_mod[1,2]
yo_se <- yo_mod[1,3]

n <- 10000
# baseline proportion simulation (using normal distribution assumptions)
yo_distribution <- rnorm(n, mean = yo_est, sd = yo_se)
# I may want to consider a beta distribution for a probability....
# luckily no values are below 0 or greater than 1, norm distribution probably okay
hist(yo_distribution)
summary(yo_distribution)

# simulating a distribution around the beta estimate (can use the logit_mod)
logit_mod
beta_est <- logit_mod[2,2]
beta_se <- logit_mod[2,3]

beta_distribution <- rnorm(n, mean = beta_est, sd = beta_se)
# distribution checks
hist(beta_distribution)
# check estimates to see if it matches with frequentist calcs of 95%CI
as.numeric(quantile(beta_distribution, 0.50, na.rm = T))
# 2.5% lower bound
as.numeric(quantile(beta_distribution, 0.025, na.rm = T))
# 97.5 % upper bound
as.numeric(quantile(beta_distribution, 0.975, na.rm = T))

# run mc estimate of delta y
# empty vector
delta_y <- vector("double", length = n)

for(i in 1:n) {
  # take one random value with replacement from distributions
  # divide by 365 to estimate daily rate
  est_y0 <- sample(yo_distribution, 1, replace = T)
  # take one random value from the beta distribution with replacement
  est_beta <- sample(beta_distribution, 1, replace = T)
  # average period estimate
  delta_y[[i]] <- est_y0*(1-exp((-est_beta)*(1)))*10000

} # end of inner loop of HIA estimate

# now lets see if the 'true' estimate is in the bounds
# median
as.numeric(quantile(delta_y, 0.50, na.rm = T))
# 2.5% lower bound
as.numeric(quantile(delta_y, 0.025, na.rm = T))
# 97.5 % upper bound
as.numeric(quantile(delta_y, 0.975, na.rm = T))

# truth 
340-161

# naw, still high 
# this is a cat
# =^o.o^=

# what if we make our confidence even higher to 99%?
# 0.5% lower bound
as.numeric(quantile(delta_y, 0.005, na.rm = T))
# 99.5 % upper bound
as.numeric(quantile(delta_y, 0.995, na.rm = T))

# a little closer

```

Well, it looks to me the HIA overestimates the 'true' count. Here it looks like we suffer from a type II error, where we missed the 'true' association. By how much, it's hard to tell since this simulated sample size is a bit small. Maybe it works better for larger populations, and small effects like air pollution. If I estimated uncertainty around delta pollutant and population at risk, that may capture the 'true' estimate.

I guess if the point of a HIA/BA is more about policy, then we'd still arrive at a conclusion that the number of outcomes that may be due to exposure is not 0. That's good I guess. 

## Estimated Poportion of Disease

What about calculating a estimated proportion of disease due to exposure? This would be the question of an 'incidence rate'. I believe this should be equal to roughly ~0.03, or attributable risk/risk-difference.

First approach. What if we just divide our estimates we obtained by the population at risk? 

$\frac{\Delta y}{pop. at risk}$

Second approach. Modify the formula. Hopefully they produce roughly the same answer.
$\frac{\Delta y}{pop. at risk} = y_{o} \times[1-e^{(-\beta \times \Delta exposure)}]$

```{r estimated proportion of disease due to exposure, message = F}

# as a reminder, here is the 'true' risk difference
# risk diff
# logistic regression beta
risk_diff_mod <- tidy(glm(disease~exposure, data = pop_data, 
                          family = "binomial"(link="identity")))
risk_diff_mod


# first approach where the bounds are divided by our population at risk (10000)
as.numeric(quantile(delta_y, 0.50, na.rm = T))/10000
# 2.5% lower bound
as.numeric(quantile(delta_y, 0.025, na.rm = T))/10000
# 97.5 % upper bound
as.numeric(quantile(delta_y, 0.975, na.rm = T))/10000
# this approach underestimates it.

# slightly modified version of this approach where I divide each estimate from the MC

# mc approach 1.1 ----
# empty vector
delta_y <- vector("double", length = n)

for(i in 1:n) {
  # take one random value with replacement from distributions
  # divide by 365 to estimate daily rate
  est_y0 <- sample(yo_distribution, 1, replace = T)
  # take one random value from the beta distribution with replacement
  est_beta <- sample(beta_distribution, 1, replace = T)
  # average period estimate
  delta_y[[i]] <- (est_y0*(1-exp((-est_beta)*(1)))*10000)/1000

} # end of inner loop of HIA estimate

# median and bounds
# median
as.numeric(quantile(delta_y, 0.50, na.rm = T))
# 2.5% lower bound
as.numeric(quantile(delta_y, 0.025, na.rm = T))
# 97.5 % upper bound
as.numeric(quantile(delta_y, 0.975, na.rm = T))

# roughly the same answer as before

# mc approach 2.0 ----
# just taking out the baseline population all together
# empty vector
delta_y <- vector("double", length = n)

for(i in 1:n) {
  # take one random value with replacement from distributions
  # divide by 365 to estimate daily rate
  est_y0 <- sample(yo_distribution, 1, replace = T)
  # take one random value from the beta distribution with replacement
  est_beta <- sample(beta_distribution, 1, replace = T)
  # average period estimate
  delta_y[[i]] <- (est_y0*(1-exp((-est_beta)*(1))))

} # end of inner loop of HIA estimate

# median and bounds
# median
as.numeric(quantile(delta_y, 0.50, na.rm = T))
# 2.5% lower bound
as.numeric(quantile(delta_y, 0.025, na.rm = T))
# 97.5 % upper bound
as.numeric(quantile(delta_y, 0.975, na.rm = T))

# same answer. Yay, math! =^O.O^=

```
